---
title: "Single Cell Analysis Workshop book 3"
output:
  html_document:
    df_print: paged
---

Here we will compare the clusters in our dataset to clusters defined by other researchers in a published datasets. Unlike methods like scClassify, MetaNeighbor does not return a label for each cell, but an overall assessment or the correspondence between pairs of clusters. We will follow this protocol: <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8826496/> (Procedure 2)

Load required libraries and set base path

```{r}
  
library(tidyverse)
library(Seurat)
library(SingleCellExperiment)
library(MetaNeighbor)
library(anndata)
library(SeuratDisk)
library(Matrix)

base_path <- "C:/Users/Home/Documents/GitHub/SingleCell_Workshop/"

```

Read in and properly format the data for the Lister lab. This is an example of using data in h5ad format with Seurat.

```{r}

#Load as anndata, and look at the object
lister_data <- read_h5ad(paste0(base_path,                      
                                "data/ListerLab/Processed_data_RNA-all_full-counts-and-downsampled-CPM.h5ad"))
lister_data

#Convert the basics to Seurat format, look at the object
lister_data <- CreateSeuratObject(counts = t(lister_data$X), meta.data = lister_data$obs)
lister_data

#Look at the major cell types and clusters, also distribution of genes and UMIs
lister_data$cell_type %>% table
lister_data$major_clust %>% table
VlnPlot(lister_data, features = c("nCount_RNA", "nFeature_RNA"))
lister_data$nCount_RNA %>% summary
lister_data$nFeature_RNA %>% summary

```

Now we will train some classifiers using this object as the reference, but first we will do some pre-processing. 

First we wiill build the classifier suing all the cells, and then we will re-run the analysis using only the non-neuronal cells. 

```{r}

#Normalize, find variable features, convert to SingleCellExperiment and subset to only the variable features 
lister_data <- subset(lister_data, cell_type == "Non-Neu")
lister_data <- NormalizeData(lister_data, normalization.method = "LogNormalize", scale.factor = 10000)
lister_data <- FindVariableFeatures(lister_data, selection.method = "vst", nfeatures = 3000)
var_features <- VariableFeatures(lister_data)
lister_data <- as.SingleCellExperiment(lister_data)
lister_data <- lister_data[var_features,]
lister_data
```

Time to train the model!

```{r}

pretrained_model_major_cluster = MetaNeighbor::trainModel(
 var_genes = rownames(lister_data),
 dat = lister_data,
 study_id = rep("Lister", dim(lister_data)[2]),
 cell_type = lister_data$major_clust
)


```

Remove the Lister lab object, and load our integrated Seurat object

```{r}
rm(lister_data)
integrated_seurat <- readRDS(paste0(base_path, "integrated_seurat.Rds"))
integrated_seurat <- as.SingleCellExperiment(integrated_seurat)

```

Classify using the major clusters and plot results

```{r}
#One versus all
aurocs = MetaNeighborUS(
 trained_model = pretrained_model_major_cluster, dat = integrated_seurat,
 study_id = rep("integrated", dim(integrated_seurat)[2]), 
 cell_type = integrated_seurat$integrated_snn_res.1.5,
 fast_version = TRUE
)
tryCatch({plotHeatmapPretrained(aurocs, margins = c(5,5))}, error = function(error_condition) {})
```

```{r}

#One versus best
best_hits = MetaNeighborUS(
 trained_model = pretrained_model_major_cluster, dat = integrated_seurat,
 study_id = rep("integrated", dim(integrated_seurat)[2]), 
 cell_type = integrated_seurat$integrated_snn_res.1.5,
 one_vs_best = TRUE,
 fast_version = TRUE
)
tryCatch({plotHeatmapPretrained(best_hits, margins = c(5,5))}, 
         error = function(error_condition) {})

```

Do some house-keeping

```{r}
print(summary(warnings()))
sessionInfo()
Sys.time()

```
