---
title: "Single Cell Analysis Workshop book 3"
output:
  html_document:
    df_print: paged
    #keep_md: yes
---

# Single Cell Analysis Workshop book 3

Here we will compare the clusters in our dataset to clusters defined by other researchers in a published datasets. Unlike methods like scClassify, MetaNeighbor does not return a label for each cell, but an overall assessment or the correspondence between pairs of clusters. We will follow this protocol: <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8826496/> (Procedure 2)

Load required libraries and set base path

```{r}
Sys.time()  
library(tidyverse)
library(Seurat)
library(SingleCellExperiment)
library(MetaNeighbor)
library(anndata)
library(SeuratDisk)
library(Matrix)
library(scclusteval)
library(mclust)

setwd("C:/Users/Home/Documents/GitHub/SingleCell_Workshop/")

```

Read in and properly format the data for the Lister lab. This is an example of using data in h5ad format with Seurat.

```{r}
#Load as anndata, and look at the object
lister_data <- read_h5ad("data/ListerLab/Processed_data_RNA-all_full-counts-and-downsampled-CPM.h5ad")
lister_data

#Convert the basics to Seurat format, look at the object
lister_data <- CreateSeuratObject(counts = t(lister_data$X), meta.data = lister_data$obs)
```


```{r}
lister_data

#Look at the major cell types and clusters, also distribution of genes and UMIs
lister_data$cell_type %>% table
lister_data$major_clust %>% table
VlnPlot(lister_data, features = c("nCount_RNA", "nFeature_RNA"), pt.size = 0.001)
lister_data$nCount_RNA %>% summary
lister_data$nFeature_RNA %>% summary

```

Now we will train some classifiers using this object as the reference, but first we will do some pre-processing. 

First we wiill build the classifier suing all the cells, and then we will re-run the analysis using only the non-neuronal cells. 

```{r}

#Normalize, find variable features, convert to SingleCellExperiment and subset to only the variable features 
lister_data <- subset(lister_data, cell_type == "Non-Neu")
lister_data <- NormalizeData(lister_data, normalization.method = "LogNormalize", scale.factor = 10000)
lister_data <- FindVariableFeatures(lister_data, selection.method = "vst", nfeatures = 3000)
var_features <- VariableFeatures(lister_data)
lister_data <- as.SingleCellExperiment(lister_data)
lister_data <- lister_data[var_features,]
lister_data
```

Time to train the model!

```{r}

pretrained_model_major_cluster = MetaNeighbor::trainModel(
 var_genes = rownames(lister_data),
 dat = lister_data,
 study_id = rep("Lister", dim(lister_data)[2]),
 cell_type = lister_data$major_clust
)


```

Remove the Lister lab object, and load our integrated Seurat object

```{r}
rm(lister_data)
integrated_seurat <- readRDS("integrated_seurat.Rds")
integrated_seurat <- as.SingleCellExperiment(integrated_seurat)

```

Classify using the major clusters and plot results

```{r}
#One versus all
aurocs = MetaNeighborUS(
 trained_model = pretrained_model_major_cluster, dat = integrated_seurat,
 study_id = rep("integrated", dim(integrated_seurat)[2]), 
 cell_type = integrated_seurat$integrated_snn_res.1.5,
 fast_version = TRUE
)
tryCatch({plotHeatmapPretrained(aurocs, margins = c(5,5))}, error = function(error_condition) {})
```

```{r}

#One versus best
best_hits = MetaNeighborUS(
 trained_model = pretrained_model_major_cluster, dat = integrated_seurat,
 study_id = rep("integrated", dim(integrated_seurat)[2]), 
 cell_type = integrated_seurat$integrated_snn_res.1.5,
 one_vs_best = TRUE,
 fast_version = TRUE
)
tryCatch({plotHeatmapPretrained(best_hits, margins = c(5,5))}, 
         error = function(error_condition) {})

```

Compare labels obtained with supervised classification of cells and unsupervised clustering, also between the two supervised clustering labels

```{r}
 integrated_seurat <- readRDS("integrated_seurat.Rds")
 plot1 <- PairWiseJaccardSetsHeatmap(Idents(integrated_seurat), integrated_seurat$Karolinski)
 adjustedRandIndex(Idents(integrated_seurat), integrated_seurat$Karolinski) %>% print
 #plot1
 plot2 <- PairWiseJaccardSetsHeatmap(Idents(integrated_seurat), integrated_seurat$Nowakowski)
 adjustedRandIndex(Idents(integrated_seurat), integrated_seurat$Nowakowski) %>% print
 #plot2
 plot3 <- PairWiseJaccardSetsHeatmap(integrated_seurat$Karolinski, integrated_seurat$Nowakowski)
 adjustedRandIndex(integrated_seurat$Karolinski, integrated_seurat$Nowakowski) %>% print
```

``` {r JaccardIndexPlots, dev = c("png")}
 #pdf(file = "C:/Users/Home/Documents/GitHub/SingleCell_Workshop/JaccardIndexPlots.pdf", onefile = TRUE)
  plot1
  plot2
  plot3  
 #dev.off()

```
Generate counts summed by or averaged by sample and cluster 

```{r}

Idents(integrated_seurat) <- paste(Idents(integrated_seurat), integrated_seurat$age, sep = ".")
aggregated_seurat <- AggregateExpression(integrated_seurat, assays = "RNA", slot = "counts", group.by = )
aggregated_seurat$RNA %>% head
aggregated_seurat$RNA %>% colSums

average_seurat <- AverageExpression(integrated_seurat, assays = "RNA", slot = "counts", group.by = )
average_seurat$RNA %>% head
average_seurat$RNA %>% colSums

```

Do some house-keeping

```{r}
print(summary(warnings()))
sessionInfo()
Sys.time()

```
