---
title: "SingleCellAnalysisWorkshop_2023 Book2"
output:
  html_document:
    df_print: paged
---

# Single Cell Analysis in R 2023 Notebook 2

This workbook presents the next steps in the basic work flow for scRNAseq analysis
1. Merging samples and batch correction
2. Clustering
3. Cluster annotation
4. Compare proportions of cell types

This workbook assumes two Seurat objects 1. adolescent and 2. adult brain samples.
The Seurat objects are created and the data is cleaned in Notebook 2. 
Running 


```{r}
Sys.time()
# load libraries
library(Seurat)
library(tidyverse)
library(enrichR)
library(ggplot2)
library("scClassify")
library(SingleCellExperiment)
library("Matrix")

# setwd only works inside the same code chunk
# we can use the GUI to change the working directory for the rest of the workbook
# Session < Set Working Directory < Choose Directory
# 

# windows
#setwd("C:/Users/Home/Documents/GitHub/SingleCell_Workshop/")

# MAC/Linux
#wd <- "/Users/rhalenathomas/Documents/scRNAclubMcGill/workshop2023/data/"
setwd("C:/Users/Home/Documents/GitHub/SingleCell_Workshop/")

# read in the data objects if starting from save files
adolescent_data_seurat <- readRDS("adolsecentSeurat.Rds") %>% RenameCells(add.cell.id = "Adol_")

adult_data_seurat <- readRDS("adultSeurat.Rds") %>% RenameCells(add.cell.id = "Adult_")

```


Merge data objects
```{r}

merge_seurat <- merge(adolescent_data_seurat,adult_data_seurat)
merge_seurat

unique(merge_seurat$orig.ident)


```

Integrate objects using the Seurat function
```{r}

sample.list <- SplitObject(merge_seurat, split.by = "orig.ident")

# We have already normalized and identified variable features in each sample
# If we had not done so we can normalize here
#for (i in 1:length(sample.list)){
 # org.list[[i]] <- NormalizeData(org.list[[i]], verbose = FALSE)
  #org.list[[i]] <- FindVariableFeatures(org.list[[i]], selection.method = "vst")
#}

# Now we find features that can act as anchors between the two samples
int.anchors <- FindIntegrationAnchors(object.list = sample.list, dims = 1:50)
integrated_seurat <- IntegrateData(anchorset = int.anchors, dims = 1:50)


```

Optional: Save the integrated object or read in the Integrated object for the next step.

```{r}
setwd(wd)

# remove # comment from the code you want to run
#saveRDS(integrated_seurat,"IntegratedSeurat.RDS")
integrated_seurat <- readRDS("IntegratedSeurat.RDS")

```


Perform dimensional reduction PCA and UMAP on the merge and integrated objects

```{r}
# first the merge object
DefaultAssay(merge_seurat) <- "RNA"

merge_seurat <- ScaleData(merge_seurat, verbose = FALSE)
# in the merge data set we sill need features for the PCA input
merge_seurat <- FindVariableFeatures(merge_seurat, selection.method = "vst")
merge_seurat <- RunPCA(merge_seurat, npcs = 30, verbose = FALSE)
merge_seurat <- RunUMAP(merge_seurat, reduction = "pca", dims = 1:30)


```


Repeat PCA and UMAP for the integrated object
```{r}
Idents(integrated_seurat) <- "integrated"
integrated_seurat <- ScaleData(integrated_seurat, verbose = FALSE)
# only the integrated features will be the pca input
integrated_seurat <- RunPCA(integrated_seurat, npcs = 30, verbose = FALSE)
integrated_seurat <- RunUMAP(integrated_seurat, reduction = "pca", dims = 1:30)

```

Optional: save the PCA and UMAP integrated object or read in the saved object
```{r}

setwd(wd)
# remove # comment from the code you want to run
saveRDS(integrated_seurat,"IntegratedSeuratGraphs.RDS")
#integrated_seurat <- readRDS("IntegratedSeuratGraphs.RDS")

```


Visualize the UMAPs comparing the merged and integrated objects

```{r}


p1 <- DimPlot(merge_seurat, group.by = "orig.ident") + ggtitle("Merge")
p2 <- DimPlot(integrated_seurat, group.by = "orig.ident") + ggtitle("Integrated")

p1
p2


```
We can see a large shift in the two populations when the data is integrated (harmonized).

Cluster the integrated data

```{r}
# see the importance of the pca components
ElbowPlot(integrated_seurat, ndims=30)
integrated_seurat

```
Choose 15 PCs
There are 9044 cells.  A common rule of thumb for choosing k for nearest neighbours is using the square root of the number of cells = 95

```{r}
# calculate the square root
sqrt(9044)

# the K parameter changes the size of clusters by changing the starting nodes input into the Louvain network

integrated_seurat <- FindNeighbors(integrated_seurat, dims = 1:15, k.param = 95)
# the number of clusters is dependent on the resolution a number from 0-2. 
# Higher values make more clusters
# we include a lot of resolutions - to save time you can just run 0.4
integrated_seurat <- FindClusters(integrated_seurat, resolution = c(0,0.05,0.25,0.4,0.5,0.6,1,1.5) )

integrated_seurat <- FindClusters(integrated_seurat, resolution = 0.4)

# we can visualize which cells are grouped together at different resolutions using clustree
library(clustree)
clustree(integrated_seurat, prefix = "integrated_snn_res.")

```

Optional: save the clusters or read in the object with clusters
```{r}
setwd(wd)
# remove # comment from the code you want to run
saveRDS(integrated_seurat,"IntegratedSeuratClusters.RDS")
#integrated_seurat <- readRDS("IntegratedSeuratClusters.RDS")

```

Visualize the UMAP of the different cluster resolutions
```{r}
res <- c(0.05,0.25,0.4,0.5,0.6,1,1.5)
resolutions <- paste("integrated_snn_res.", res, sep="")
resolutions

for(r in resolutions){
  print(DimPlot(integrated_seurat, group.by = r))
}


```

Now we choose a resolution to annotate. We are expecting microglia, astrocytes, oligodendrocytes and OPCs. Resolution 0.4 looks like a good amount of clusters.


Cluster annotation


```{r}
#setwd(wd)

# find cluster markers

# we need to select the level of clustering that we want to annotate
Idents(integrated_seurat) <- "integrated_snn_res.0.4"
ClusterMarkers <- FindAllMarkers(integrated_seurat, only.pos = TRUE)

# optional: save the cluster markers
write.csv(ClusterMarkers,"ClusterMarkers.csv")

```

Optional: If cluster markers were run previously and saved or if they take too long to run, read in the csv
```{r}

#setwd(wd)
ClusterMarkers <- read.csv("ClusterMarkers.csv")

  
```


Lets look at the markers to see if they identify cell types

```{r}
head(ClusterMarkers)
top5 <- ClusterMarkers %>% group_by(cluster) %>% top_n(n=5, wt =avg_log2FC)
DoHeatmap(integrated_seurat, features = top5$gene, size = 3, angle = 90, group.by = "integrated_snn_res.0.4")

```
We can see cluster 1 and have a huge overlap. Cluster 5 and 6 are also overlapping.


We can look at the markers for each cluster using Gene set analysis or cell type libraries

```{r}

setEnrichrSite("Enrichr") # Human genes
# list of all the databases

dbs <- listEnrichrDbs()

# this will list the possible libraries
dbs

# select libraries with cell types
db <- c('CellMarker_Augmented_2021','Azimuth_Cell_Types_2021')




```

Here is a small function to run easily on each cluster and find the cell type library predictions

```{r}

checkCelltypes <- function(cluster_num = 0){
  clusterX <- ClusterMarkers %>% filter(cluster == cluster_num & avg_log2FC > 0.25)
  genes <- clusterX$gene
  # the cell type libraries
  # get the results for each library
  clusterX.cell <- enrichr(genes, databases = db)
  # visulize the results
print(plotEnrich(clusterX.cell[[1]], showTerms = 20, numChar = 40, y = "Count", orderBy = "P.value", title = 'CellMarker_Augmented_2021'))
print(plotEnrich(clusterX.cell[[2]], showTerms = 20, numChar = 40, y = "Count", orderBy = "P.value", title = 'Azimuth_Cell_Types_2021'))

}



```

Run the function for each cluster to see if we can identify cell types

```{r}

cluster0 <- checkCelltypes(cluster_num = 0)



```
Cluster 0 is some kind of immune cells, possibly dendritic cell or macrophage.  These are likely the microglia

```{r}
cluster1 <- checkCelltypes(cluster_num = 1)
```
Cluster 1 appears to have oligodendrocytes

```{r}
cluster2 <- checkCelltypes(cluster_num = 2)
```
Cluster 2 is also immune cells, embryonic microglia.

```{r}
# cluster 3
cluster3 <- checkCelltypes(cluster_num = 3)

```
Cluster 3 is oligodendrocytes

```{r}
# cluster 4
cluster4 <- checkCelltypes(cluster_num = 4)
```
Cluster 4 - immune cells


```{r}
# cluster 5
cluster5 <- checkCelltypes(cluster_num = 5)

```
Cluster 5 is astrocytes


```{r}
# cluster 6
cluster6 <- checkCelltypes(cluster_num = 6)

```
Cluster 6 is OPCs

```{r}
# cluster 7
cluster7 <- checkCelltypes(cluster_num = 7)

```
Cluster 7 - monocytes or Tcells

We can remove the lists of cell library outputs to clean up our work environment
```{r}
rm(cluster0,cluster1,cluster2,cluster3,cluster4,cluster5,cluster6,cluster7)

rm(adolescent_data_seurat, adult_data_seurat, bcmvn_adolescent, bcmvn_adult, int.anchors, merge_seurat, p1, p2, sample.int, sample.list, sweep.res.list_adolescent, sweep.res.list_adult, sweep.stats_adolescent, sweep.stats_adult)
# don't worry about the warning

```


Looking at known cell type markers is another way to identify cell types
```{r}
# marker lists

#  doi: 10.1002/glia.23767
microglia <- c("IBA1","P2RY12","P2RY13","TREM119", "GPR34","SIGLECH","TREM2",
               "CX3CR1","FCRLS","OLFML3","HEXB","TGFBR1", "SALL1","MERTK",
               "PROS1")


oligodendrocyte <- c("MBP","MOG","OLIG1","OLIG2","SOX10")

# doi: 10.3390/biom11091361
astrocytes <- c("GFAP","S100B","AQP4","APOE", "SOX9")


integrated_seurat <- AddModuleScore(integrated_seurat, features = list(astrocytes, oligodendrocyte), name = "Glia")

VlnPlot(integrated_seurat, features = c("Glia1", "Glia2"))

```


Look at the expression of different marker lists
```{r}
# we want to use the total RNA not just the genes that are integrated
# change the default assay
DefaultAssay(integrated_seurat) <- "RNA"
# we need to scale the RNA data
integrated_seurat <- ScaleData(integrated_seurat)
DoHeatmap(integrated_seurat, features = microglia, group.by = "integrated_snn_res.0.4")
DotPlot(integrated_seurat, features = microglia, group.by = "integrated_snn_res.0.4") + RotatedAxis()
Idents(integrated_seurat) <- "integrated_snn_res.0.4"
FeaturePlot(integrated_seurat, features = microglia, label = TRUE)

```
Cluster 0 has some microglial markers and cluster 2 has lots of microglia markers.


```{r}
# oligodendrocytes
DoHeatmap(integrated_seurat, features = oligodendrocyte, group.by = "integrated_snn_res.0.4")
DotPlot(integrated_seurat, features = oligodendrocyte, group.by = "integrated_snn_res.0.4") + RotatedAxis()

```
Cluster 1,3,5 have mature oligodendrocyte markers and cluster 6 has strong oligo lineage markers 

```{r}
# astrocyte
DoHeatmap(integrated_seurat, features = astrocytes, group.by = "integrated_snn_res.0.4")
DotPlot(integrated_seurat, features = astrocytes, group.by = "integrated_snn_res.0.4") + RotatedAxis()

```
Cluster 5 is the best match for astrocyte markers

Optional save and read in complete object

```{r}
setwd(wd)
saveRDS(integrated_seurat, file = "integrated_seurat.Rds")
integrated_seurat <- readRDS("integrated_seurat.Rds")

```

Automated cell type annotation

scClassify R weighted kNN https://doi.org/10.15252/msb.20199389

https://sydneybiox.github.io/scClassify/articles/scClassify.html

```{r}

# we need to create the reference matrices
#matrices where each row is a gene and each column a cell for a reference dataset and a query data set

# reference data
# we have the data prepared as a Seurat object

ref <- readRDS("/Users/rhalenathomas/Documents/Data/scRNAseq/PublicData/Karolinski_DevForebrain_downsample_Level1.RDS")
ref

#check the meta data
colnames(ref@meta.data)

unique(ref$Clusters)
unique(ref$Level1)
unique(ref$Celltypes)

# make the identity the Cell types 
Idents(ref) <- "Celltypes"
dim(ref)


```
The labelled cell types are in WGCNAcluster but there are too many to work well as a reference matrix.  scClassify suggests recommends to start with fewer cell types.


We need to make the reference and query objects into a "dgCMatrix" object

```{r}


# make a dfCMatrix from Seurat object to use in scClassify

dgCMat_ref <- GetAssayData(ref)
# get the cell type data
Idents(ref) <- "Celltypes"
ref_celltypes <- Idents(ref)
unique(ref_celltypes)


# convert the integrated clustered object into dgCMatrix
DefaultAssay(integrated_seurat) <- "RNA"
dgCMat_query <- GetAssayData(integrated_seurat)
Idents(integrated_seurat) <- "Original_clusterID"
query_celltypes <- Idents(integrated_seurat)

# check the objects
dim(dgCMat_ref)
dim(dgCMat_query)

class(dgCMat_ref)
class(dgCMat_query)

```

See the original cell type annotations
```{r}
table(ref_celltypes)
table(query_celltypes)

length(ref_celltypes)
length(query_celltypes)
```


We can now perform non-ensemble scClassify

```{r}
scClassify_res <- scClassify(exprsMat_train = dgCMat_ref,
                             cellTypes_train = ref_celltypes,
                             exprsMat_test = dgCMat_query,
                             cellTypes_test = query_celltypes,
                             tree = "HOPACH",
                             algorithm = "WKNN",
                             selectFeatures = c("limma"),
                             similarity = c("pearson"),
                             returnList = FALSE,
                             verbose = FALSE)
```

We can check the cell type tree generated by the reference data:

```{r}
scClassify_res$trainRes
plotCellTypeTree(cellTypeTree(scClassify_res$trainRes))

```

Check the predictions
```{r}

table(scClassify_res$testRes$test$pearson_WKNN_limma$predRes,query_celltypes)



```

We want to predict cluster labels

```{r}

# we need to make the test meta data into the cluster ids 
Idents(integrated_seurat) <- "integrated_snn_res.0.4"
query_clusters <- Idents(integrated_seurat)


scClassify_res <- scClassify(exprsMat_train = dgCMat_ref,
                             cellTypes_train = ref_celltypes,
                             exprsMat_test = dgCMat_query,
                             cellTypes_test = query_clusters,
                             tree = "HOPACH",
                             algorithm = "WKNN",
                             selectFeatures = c("limma"),
                             similarity = c("pearson"),
                             returnList = FALSE,
                             verbose = FALSE)
```


See the cell type predictions by cluster and add the predicted labels to the Seurat object

```{r}

table(scClassify_res$testRes$test$pearson_WKNN_limma$predRes,query_clusters)
meta_to_add <- scClassify_res$testRes$test$pearson_WKNN_limma$predRes %>% 
                enframe %>% 
                as.data.frame %>% 
                column_to_rownames("name") %>%
                setNames(nm = c("Karolinski"))
integrated_seurat <- AddMetaData(integrated_seurat, meta_to_add)

```

Check another reference set

```{r}
# refrence data
# we have the data prepared as a Seurat object

ref2 <- readRDS("/Users/rhalenathomas/Documents/Data/scRNAseq/PublicData/Nowakowski_dev_cortext.RDS")
ref2

#check the meta data
colnames(ref2@meta.data)

#make the reference matrix and reference cell types 

dgCMat_ref2 <- GetAssayData(ref2)
# get the cell type data
Idents(ref2) <- "Celltypes"

levels(ref2)

ref_celltypes2 <- as.character(Idents(ref2))
ref_celltypes2[is.na(ref_celltypes2)] <- "NA"

unique(ref_celltypes2)

# check the object
dim(dgCMat_ref2)


class(dgCMat_ref2)



```

```{r}
# predict the cluster cell types from the second reference
scClassify_res <- scClassify(exprsMat_train = dgCMat_ref2,
                             cellTypes_train = ref_celltypes2,
                             exprsMat_test = dgCMat_query,
                             cellTypes_test = query_clusters,
                             tree = "HOPACH",
                             algorithm = "WKNN",
                             selectFeatures = c("limma"),
                             similarity = c("pearson"),
                             returnList = FALSE,
                             verbose = FALSE)

#Add the predicted labels to the Seurat object 

meta_to_add <- scClassify_res$testRes$test$pearson_WKNN_limma$predRes %>% 
                enframe %>% 
                as.data.frame %>% 
                column_to_rownames("name") %>%
                setNames(nm = c("Nowakowski"))
integrated_seurat <- AddMetaData(integrated_seurat, meta_to_add)

```

Now that we have the different forms of information to annotate the cells we need to decide on the annotations to add, then add them into the data object.

```{r}
# we need to set the identity to rename
Idents(integrated_seurat) <- "integrated_snn_res.0.4"

# we need to make a vector of the cell type in the same order - in the cluster order

cell_types <- c("microglia","oligodendrocytes_1","embyonic_microglia",
                "oligodendrocyte_2","immune", "astrocytes","OPC","monocytes")
  
names(cell_types) <- levels(integrated_seurat)
integrated_seurat <- RenameIdents(integrated_seurat, cell_types)
integrated_seurat <- AddMetaData(object=integrated_seurat, metadata=Idents(integrated_seurat), col.name = "CellTypes")

```

Save the integrated object for next steps

```{r}
setwd(wd)
saveRDS(integrated_seurat, file = "integrated_seurat.Rds")

Sys.time()

```



